---
//title: "Murders-analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of murders dataset
Dataset was provided as part of the edX course HarvardX PH125.1x: Data Science: R Basics </BR>
The coursework was originally done through DataCamp. </BR>
Redid the work using R Studio as a practice exercise.

### Load in dataset

```{r dataset}
library(tidyverse)
murders <- read_csv("./data/murders.csv")
names(murders)
```

Look at dataset content
```{r}
str(murders)
```

Look at first 6 records
```{r}
head(murders)
```

Find the state with the lowest total number of murders
```{r}
murders$state[which.min(murders$total)]
```

Find the state with the highest total number of murders
```{r}
murders$state[which.max(murders$total)]
```

Create a vector for the state names </BR>
Order the states by population
```{r}
states <- murders$state
ranks <- rank(murders$population)
index <- order(murders$population)
states_in_order <- data.frame( state_name = states[index],
                               state_rank = ranks[index])
states_in_order
```
Calculate the murder rate per capita </BR>
Order the states by descending murder rate
```{r}
murder_rate <- murders$total / murders$population * 100000
murders$state[ order(murder_rate, decreasing = TRUE)]
```
We want to find the states where the murder rate is less than 0.71
```{r}
sum(murder_rate < 0.71)
murders$state[murder_rate < 0.71]
```
Find states that meet two conditions: Region = West and murder rate is less than 1
```{r}
west <- murders$region == "West"
safe <- murder_rate <=1
index <- safe & west
murders$state[index]
```
Find the murder rate for Massachusetts
```{r}
murder_rate[which(murders$state == "Massachusetts")]
```

## dplyr package
Install the dplyr package
```{r}
library(dplyr)
```

Add a new column to the table, for the murder rate per capita
```{r}
murders <- mutate(murders, rate = total / population * 100000)
head(murders)
```
Display the 5 states with a murder rate less than 0.71
```{r}
filter(murders, rate < 0.71)
```
Display only 3 columns for the 5 states with the lowest rate
```{r}
murders %>% select(state, region, rate) %>% filter( rate < 0.71 )
```
Create a scatterplot of population in millions & total murders
```{r}
population_in_millions <- murders$population / 1000000
total_gun_murders <- murders$total
plot(population_in_millions, total_gun_murders)
```
</BR>
The graph shows that there is a relationship between population and number of murders </BR>

Create a histogram for the murder rate
```{r}
hist(murders$rate)
```
</BR>
There is one very high anomaly
```{r}
murders$state[which.max(murders$rate)]
```
Create a boxplot
```{r}
boxplot(rate~region, data = murders)
```
</BR>
The region South has the highest murder rate </BR>

